\stktokens{}: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities\\ Technical Report with Proves and Details

### 2018-07-05 fully-abstract-comp/technical-report.tex
Small variable name correction
Added a todo in tech report.
Remark about missing gc in source instruction interpretation.

We use the following step relation:
\lau{RESOLVED: Add $\ta$ to the source step.}
\lau{RESOLVED: Add $\stkb$ to source term and everything else where it is necessary. Consider how $\stkb$ and $\ta$ can be added without passing around these values in LR and source semantics (solution to parameterize both with it, so they need to be provided only once.)}
\begin{align*}
  \src{\Phi} & \; \src{\step[\src{\ta,\stkb}] \sem{\scall{\offpc,\offsigma}{r_1}{r_2}}(\Phi)} &  &\sourcecolor\left.\arraycolsep=0pt
                                                  \begin{array}[t]{l}
                                                    \text{if }\Phi(\pcreg) 
= ((\perm,\_),\baddr,\eaddr,\aaddr) \tand \\
                                                    \callCond{\Phi,r_1,r_2,\offpc,\offsigma,\aaddr} \tand\\
                                                    \lbrack \aaddr ,\aaddr + \calllen - 1\rbrack \subseteq \ta \tand \\
                                                    \lbrack\aaddr,\aaddr + \calllen-1\rbrack \subseteq [\baddr,\eaddr] \tand \\
                                                    \exec{\Phi(\pcreg)}
                                                  \end{array}\right.\\
  \Phi & \step[\src{\ta,\stkb}] \sem{\decInstr{\Phi.\mem(\aaddr)}}(\Phi) & &\left.\arraycolsep=0pt
                                                  \begin{array}[t]{l}
                                                    \text{if }\Phi(\pcreg) = ((\perm,\_),\baddr,\eaddr,\aaddr) \tand \\
                                                    \src{(\neg\callCond{\Phi,r_1,r_2,\offpc,\offsigma,\aaddr} \tor}\\
                                                    \src{\eaddr < \aaddr + \calllen-1) \tand} \\
                                                    \withinBounds{\Phi(\pcreg)} \tand \\
                                                    \exec{\Phi(\pcreg)}
                                                  \end{array}\right.\\
  \Phi& \step[\src{\ta,\stkb}] \failed & & \totherwise
\end{align*}
\dominique{Note: The above does not allow executing code that is on the stack in the source machine, while this is in principle allowed in the target.
This is probably fine if we make sure a stack pointer can never be executable.}

@@

Two important points about $\tsplice{}{}{}$ related to the calling convention: (1) Splice fails if two capabilities are not adjacent. This means that if a caller tries to use a return pointer with a stack that is not immediately adjacent to the private stack, then it fails. (2) Splice prohibit splicing with an empty capability! This means that a callee cannot return an empty stack (this also means that it is impossible to make a call when all of the stack is used - this may indeed be undesirable, but without this restriction we need to handle other things).
Note: because $\tsplice{}{}{}$ does not allow empty stacks, it is not ``left inverse'' to $\tsplit{}{}{}{}$ (because of the empty case). Intuitively, it is weird that a $\tsplit{}{}{}{}$ followed by a $\tsplice{}{}{}$ does not yield the same capability.

@@

We define a component as follows:
\begin{align*}
  s &\in \Symbol\\
  \var{import} &\mathrel{::=} a \mapsfrom s\\
  \var{export} &\mathrel{::=} s \mapsto w\\
  \var{comp}_0 &\mathrel{::=} (\mscode,\msdata,\overline{c_{\mathrm{import}}},\overline{c_{\mathrm{export}}},\sigrets,\sigcloss,A_\linear)\\
  \var{comp} &\mathrel{::=} \var{comp}_0\\
               &\mid  (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})
\end{align*}
\clearpage
We define inductively when a component is valid ($\vdash \var{comp}$) by the below inference rules:
\begin{mathpar}
  \inference{
    \mscode(a) = \seal{\sigma_\baddr,\sigma_\eaddr,\sigma_\baddr} & [\sigma_\baddr,\sigma_\eaddr] = (\sigrets \cup \sigcloss)
  }{
    \sigrets,\sigrets[\mathrm{owned}],\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a
  }
  \and
  \inference{
    \mscode(a) \in \ints\\
    \mscode([a \cdots a + \calllen-1]) = \scall[0..\calllen-1]{\offpc,\offsigma}{r_1}{r_2} \Rightarrow (\mscode(a+\offpc) = \seal{\sigma_\baddr,\sigma_\eaddr,\sigma_\baddr} \wedge \sigma_\baddr+\offsigma \in \sigrets[\mathrm{owned}])
  }{
    \sigrets,\sigrets[\mathrm{owned}],\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a
  }
  \and
  \inference{
    \text{$\mscode$ has no hidden calls}\\
    \sigrets \mathrel{\#} \sigcloss \\
    \exists d_\sigma : \dom(\mscode) \rightarrow \powerset{\Seal} \ldotp
    \sigrets = \biguplus_{a \in \dom(\mscode)} d_\sigma(a) \tand\\
    \forall a \in \dom(\mscode) \ldotp 
    \sigrets,d_\sigma(a),\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a\\
    \exists a \ldotp \mscode(a) = \seal{\sigma_\baddr,\sigma_\eaddr,\_} \wedge [\sigma_\baddr,\sigma_\eaddr] \neq \emptyset
  }{
    \sigrets,\sigcloss \vdash_{\mathrm{comp-code}} \mscode
  }
\end{mathpar}

@@

\begin{mathpar}
  \inference{
    \var{comp}_1 = (\mscode[1], \msdata[1], \overline{c_{\mathrm{import},1}}, \overline{c_{\mathrm{export},1}}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \var{comp}_2 = (\mscode[2], \msdata[2], \overline{c_{\mathrm{import},2}}, \overline{c_{\mathrm{export},2}}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \var{comp}_3 = (\mscode[3], \msdata[3], \overline{c_{\mathrm{import},3}}, \overline{c_{\mathrm{export},3}}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \var{comp}_3 = \var{comp}_1 \bowtie \var{comp}_2
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \var{comp}_0 \bowtie (\var{comp}_0',c_{\mathrm{main},c}, c_{\mathrm{main},d})
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \var{comp}_0'
  }
\end{mathpar}

>>>>

We use the following step relation:
\lau{RESOLVED: Add $\ta$ to the source step.}
\lau{RESOLVED: Add $\stkb$ to source term and everything else where it is necessary. Consider how $\stkb$ and $\ta$ can be added without passing around these values in LR and source semantics (solution to parameterize both with it, so they need to be provided only once.)}
\begin{align*}
  \src{\Phi} & \; \src{\step[\src{\ta,\stkb}] \sem{\scall{\offpc,\offsigma}{r_1}{r_2}}(\Phi)} &  &\sourcecolor\left.\arraycolsep=0pt
                                                  \begin{array}[t]{l}
                                                    \text{if }\Phi(\pcreg) 
= ((\perm,\_),\baddr,\eaddr,\aaddr) \tand \\
                                                    \callCond{\Phi,r_1,r_2,\offpc,\offsigma,\aaddr} \tand\\
                                                    \lbrack \aaddr ,\aaddr + \calllen - 1\rbrack \subseteq \ta \tand \\
                                                    \lbrack\aaddr,\aaddr + \calllen-1\rbrack \subseteq [\baddr,\eaddr] \tand \\
                                                    \exec{\Phi(\pcreg)}
                                                  \end{array}\right.\\
  \Phi & \step[\src{\ta,\stkb}] \sem{\decInstr{\Phi.\mem(\aaddr)}}(\Phi) & &\left.\arraycolsep=0pt
                                                  \begin{array}[t]{l}
                                                    \text{if }\Phi(\pcreg) = ((\perm,\_),\baddr,\eaddr,\aaddr) \tand \\
                                                    \src{(\neg\callCond{\Phi,r_1,r_2,\offpc,\offsigma,\aaddr} \tor}\\
                                                    \src{\eaddr < \aaddr + \calllen-1) \tand} \\
                                                    \withinBounds{\Phi(\pcreg)} \tand \\
                                                    \exec{\Phi(\pcreg)}
                                                  \end{array}\right.\\
  \Phi& \step[\src{\ta,\stkb}] \failed & & \totherwise
\end{align*}
On the source machine, the instruction interpretation also takes $\gc$, the global constants.
It does, however, just pass it around and never makes changes to it, so we will leave it implicit.

\dominique{Note: The above does not allow executing code that is on the stack in the source machine, while this is in principle allowed in the target.
This is probably fine if we make sure a stack pointer can never be executable.}

@@

\lau{(important) TODO 05-07-2018: It seems that there may be an issue related to splicing stk-ptr with a linear capability.
  This would be okay on trg machine, but fail on src machine.
  It may possibly be fixed simply by padding the stack with un-addressable memory (like we did with code memory).}
Two important points about $\tsplice{}{}{}$ related to the calling convention: (1) Splice fails if two capabilities are not adjacent.
This means that if a caller tries to use a return pointer with a stack that is not immediately adjacent to the private stack, then it fails.
(2) Splice prohibit splicing with an empty capability!
This means that a callee cannot return an empty stack (this also means that it is impossible to make a call when all of the stack is used - this may indeed be undesirable, but without this restriction we need to handle other things).
Note: because $\tsplice{}{}{}$ does not allow empty stacks, it is not ``left inverse'' to $\tsplit{}{}{}{}$ (because of the empty case).
Intuitively, it is weird that a $\tsplit{}{}{}{}$ followed by a $\tsplice{}{}{}$ does not yield the same capability.

@@

We define a component as follows:
\begin{align*}
  s &\in \Symbol\\
  \var{import} &\mathrel{::=} a \mapsfrom s\\
  \var{export} &\mathrel{::=} s \mapsto w\\
  \var{comp}_0 &\mathrel{::=} (\mscode,\msdata,\overline{\var{import}},\overline{\var{export}},\sigrets,\sigcloss,A_\linear)\\
  \var{comp} &\mathrel{::=} \var{comp}_0\\
               &\mid  (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})
\end{align*}
\clearpage
We define inductively when a component is valid ($\vdash \var{comp}$) by the below inference rules:
\begin{mathpar}
  \inference{
    \mscode(a) = \seal{\sigma_\baddr,\sigma_\eaddr,\sigma_\baddr} & [\sigma_\baddr,\sigma_\eaddr] = (\sigrets \cup \sigcloss)
  }{
    \sigrets,\sigrets[\mathrm{owned}],\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a
  }
  \and
  \inference{
    \mscode(a) \in \ints\\
    \mscode([a \cdots a + \calllen-1]) = \scall[0..\calllen-1]{\offpc,\offsigma}{r_1}{r_2} \Rightarrow (\mscode(a+\offpc) = \seal{\sigma_\baddr,\sigma_\eaddr,\sigma_\baddr} \wedge \sigma_\baddr+\offsigma \in \sigrets[\mathrm{owned}])
  }{
    \sigrets,\sigrets[\mathrm{owned}],\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a
  }
  \and
  \inference{
    \text{$\mscode$ has no hidden calls}\\
    \sigrets \mathrel{\#} \sigcloss \\
    \exists d_\sigma : \dom(\mscode) \rightarrow \powerset{\Seal} \ldotp
    \sigrets = \biguplus_{a \in \dom(\mscode)} d_\sigma(a) \tand\\
    \forall a \in \dom(\mscode) \ldotp 
    \sigrets,d_\sigma(a),\sigcloss \vdash_{\mathrm{comp-code}} \mscode,a\\
    \exists a \ldotp \mscode(a) = \seal{\sigma_\baddr,\sigma_\eaddr,\_} \wedge [\sigma_\baddr,\sigma_\eaddr] \neq \emptyset
  }{
    \sigrets,\sigcloss \vdash_{\mathrm{comp-code}} \mscode
  }
\end{mathpar}

@@

\begin{mathpar}
  \inference{
    \var{comp}_1 = (\mscode[1], \msdata[1], \overline{\var{import}_1}, \overline{\var{export}_1}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \var{comp}_2 = (\mscode[2], \msdata[2], \overline{\var{import}_2}, \overline{\var{export}_2}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \var{comp}_3 = (\mscode[3], \msdata[3], \overline{\var{import}_3}, \overline{\var{export}_3}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \var{comp}_3 = \var{comp}_1 \bowtie \var{comp}_2
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \var{comp}_0 \bowtie (\var{comp}_0',c_{\mathrm{main},c}, c_{\mathrm{main},d})
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \var{comp}_0'
  }
\end{mathpar}
